---
title: "CoDesymNMF_example"
author: "Annie Xie"
date: "2025-03-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
This is an example of how to run coordinate descent symNMF (CoDesymNMF).

```{r}
library(ggplot2)
library(pheatmap)
```

```{r}
source('code/coordinate_descent_symNMF.R')
source('code/visualization_functions.R')
```

# Example

## Data Generation
In this section, we simulate tree-structured data with 4 populations.

```{r}
sim_4pops <- function(args) {
  set.seed(args$seed)
  
  n <- sum(args$pop_sizes)
  p <- args$n_genes
  
  FF <- matrix(rnorm(7 * p, sd = rep(args$branch_sds, each = p)), ncol = 7)
  # if (args$constrain_F) {
  #   FF_svd <- svd(FF)
  #   FF <- FF_svd$u
  #   FF <- t(t(FF) * branch_sds * sqrt(p))
  # }
  
  LL <- matrix(0, nrow = n, ncol = 7)
  LL[, 1] <- 1
  LL[, 2] <- rep(c(1, 1, 0, 0), times = args$pop_sizes)
  LL[, 3] <- rep(c(0, 0, 1, 1), times = args$pop_sizes)
  LL[, 4] <- rep(c(1, 0, 0, 0), times = args$pop_sizes)
  LL[, 5] <- rep(c(0, 1, 0, 0), times = args$pop_sizes)
  LL[, 6] <- rep(c(0, 0, 1, 0), times = args$pop_sizes)
  LL[, 7] <- rep(c(0, 0, 0, 1), times = args$pop_sizes)
  
  E <- matrix(rnorm(n * p, sd = args$indiv_sd), nrow = n)
  Y <- LL %*% t(FF) + E
  YYt <- (1/p)*tcrossprod(Y)
  return(list(Y = Y, YYt = YYt, LL = LL, FF = FF, K = ncol(LL)))
}
```

```{r}
sim_args_4pop <- list(seed = 1, pop_sizes = rep(40,4), n_genes = 1000, branch_sds = rep(2,7), indiv_sd = 1)
baltree_4pop_1 <- sim_4pops(sim_args_4pop)
```

This is a heatmap of the true loadings matrix:
```{r}
plot_heatmap(baltree_4pop_1$LL)
```

## Running CoDesymNMF

The CoDesymNMF method aims to minimize the following objective function:
$$\|A - HH'\|_F^2 $$
where $A$ is a $n \times n$ non-negative, symmetric matrix and $H$ is a $n \times r$ non-negative matrix.

The function has 5 arguments. The first, `A` is the symmetric matrix you wish to factorize. The second, `r`, is the number of columns you want $H$ to have. The third, `init_type`, is the type of initialization you want to use. The two options are `zeros`, which initializes $H$ to a matrix of zeroes and `random`, which initializes $H$ to a randomly generated $n \times r$ matrix. The fourth input, `maxiter` is the number of iterations you want to run. The last input, `rand_permutation`, is a TRUE/FALSE value indicating whether you want to randomly permute the columns of $H$. The paper mentions that when using random initialization, randomly shuffling the columns generally performs better during each iteration (An iteration involves one pass through all the columns of $H$).

The output of the function is the estimate for $H$ and the final objective function value.

```{r}
CoDesymNMF_fit_obj <- CoDesymNMF(A = baltree_4pop_1$YYt, r = 7, init_type = 'zeros', maxiter = 100, rand_permutation = FALSE)
```

## Visualizing Results

This is a heatmap of the estimate for $H$ from CoDesymNMF:
```{r}
plot_heatmap(CoDesymNMF_fit_obj$H)
```

This is a scatter plot of the estimate for $H$ from CoDesymNMF:
```{r}
pops <- c(rep('A', 40), rep('B', 40), rep('C', 40), rep('D', 40))
plot_loadings(CoDesymNMF_fit_obj$H, pops)
```

